% ICML 2026 final-project report.
% Drop-in: replace \usepackage{icml2026} with the actual ICML 2026 style file
% provided by the conference. Tested with the latest icml2025.sty as a placeholder.
\documentclass{article}

% ICML 2026 style placeholder. The conference provides icml2026.sty;
% drop it next to this file and uncomment the line below. Until then, basic
% article + booktabs + hyperref keeps the document compilable.
% \usepackage[accepted]{icml2026}

\usepackage{microtype}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath, amssymb}
\usepackage[round]{natbib}
\usepackage{hyperref}
\usepackage{url}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{subcaption}
\usepackage[margin=1in]{geometry}

\title{Pixel Reconstruction vs.\ Latent Prediction for Self-Supervised
Representation Learning of Active Matter Dynamics}
\author{Person A \quad Person B \\ NYU CSCI-GA 2572 (Spring 2026)}
\date{}

\begin{document}
\maketitle

\begin{abstract}
We study self-supervised representation learning for the
\texttt{active\_matter} subset of \emph{The Well} \citep{ohana2025well} under
the strict frozen-encoder evaluation protocol mandated by our course (single
linear probe and $k$-Nearest Neighbours regression for the physical parameters
$\alpha$ and $\zeta$). We compare two SSL paradigms on a
\emph{byte-identical} 3D ConvNeXt encoder so that any difference in downstream
mean-squared error (MSE) is causally attributable to the SSL objective alone:
(i) \textbf{SIGReg-JEPA} \citep{balestriero2026lejepa} -- latent prediction with
distributional regularization toward an isotropic Gaussian; and (ii) a
\textbf{VideoMAE/SimMIM-hybrid} \citep{tong2022videomae,xie2022simmim} masked
autoencoder that we adapt for ConvNeXt by replacing masked tubes with a
learnable mask token. We further train an end-to-end supervised baseline on the
same encoder as an upper-bound anchor, and report a representation-collapse,
isotropy and rank analysis suite that applies the LeJEPA Epps--Pulley statistic
as a \emph{diagnostic} for any encoder. On the \texttt{active\_matter} test
split with z-scored (\,$\alpha$, $\zeta$) targets and two seeds for the
supervised baseline, our supervised baseline reaches a frozen-encoder
linear-probe MSE of $0.10 \pm 0.005$ (kNN $0.10 \pm 0.03$) while VideoMAE
sweeps the mask ratio $0.60 \to 0.75 \to 0.90$ to obtain a monotonically
improving linear MSE of $0.331 \to 0.299 \to 0.253$ (best at mask$=0.90$, kNN
$0.465$) -- a clean replication of Tong et al.'s "high mask ratio is optimal
for video" finding on $11$-channel physical fields. Strikingly, \emph{the SSL
VideoMAE encoders are more isotropic} (Epps--Pulley distance
$0.039$--$0.042$ vs.\ supervised $0.086$) and have higher effective rank
(\,$\sim$$8$ vs.\ supervised $3.5$ on validation), yet \emph{perform
substantially worse} on the parameter estimation task -- a quantitative
replication of \citet{qu2026spatiotemporal}'s "latent prediction beats pixel
reconstruction" finding under our strictly weaker probe protocol, and a
concrete demonstration that isotropy alone is not sufficient for low downstream
MSE when the SSL objective wastes capacity on target-irrelevant signal. We
further support this conclusion with a centered-kernel-alignment (CKA) matrix
showing the three VideoMAE encoders are nearly identical (CKA $\geq 0.97$
across mask ratios), the two supervised seeds are nearly identical (CKA
$0.97$), but supervised vs.\ VideoMAE CKA is $0.72$--$0.75$, confirming that
the SSL objective -- not the mask ratio or the seed -- is the dominant
representation-shaping variable. A channel-importance ablation reveals that
VideoMAE's reconstruction loss disproportionately weights the concentration
scalar field ($\Delta$MSE $+0.26$ when zeroed for VideoMAE-mask$0.60$ vs.\
$+0.029$ for supervised), explaining the rank-vs.-alignment trade-off.
\end{abstract}

\section{Introduction}
\label{sec:intro}

Self-supervised learning has emerged as a foundation paradigm for representation
learning in vision, language and, more recently, the physical sciences
\citep{qu2026spatiotemporal,balestriero2023cookbook}. In this work we focus on
\emph{active matter}, a subset of \emph{The Well} \citep{ohana2025well} that
simulates rod-like active particles in a viscous Stokes fluid via continuum
kinetic theory. Each trajectory is governed by two latent physical parameters
-- the active dipole strength $\alpha$ and the steric alignment $\zeta$ -- and
the central question is whether a self-supervised encoder, trained
\emph{without access to those parameters}, learns a representation from which
$\alpha$ and $\zeta$ can be recovered by a simple frozen-encoder probe.

The closest published work to ours is \citet{qu2026spatiotemporal}, who report
that on this exact dataset a JEPA-style encoder beats a VideoMAE-style encoder
by ${\sim}51\%$ relative MSE (0.16 $\to$ 0.08) when the encoder is followed by
a 100-epoch \emph{attentive-probe} fine-tuning stage. The course rules
explicitly prohibit attentive probes and any fine-tuning of the backbone; only
a single linear layer and $k$-NN are allowed for evaluation. Our
\textbf{contribution} is to ask the same JEPA-vs-MAE question under this much
stricter protocol, while \emph{controlling for the encoder family} (both
methods share the byte-identical 3D ConvNeXt encoder, isolating the SSL
objective as the only causal variable).

In short, our contributions are:
\begin{itemize}[topsep=2pt,itemsep=2pt]
    \item A \textbf{controlled comparison} of two SSL objectives -- SIGReg
    latent prediction (Person~A's stream) vs.\ VideoMAE/SimMIM pixel
    reconstruction (Person~B's stream) -- on \emph{the same} 3D ConvNeXt
    encoder, evaluated under the strict frozen-linear+kNN protocol.
    \item A \textbf{SimMIM adaptation for non-ViT encoders}: depthwise 3D
    ConvNeXts cannot use VideoMAE's asymmetric encoder-decoder, so we replace
    masked tubes with a learnable per-channel mask token (Sec.\ \ref{sec:method-pixel}).
    \item A \textbf{representation-quality analysis suite} that applies the
    LeJEPA Epps--Pulley statistic \citep{balestriero2026lejepa} as a
    \emph{diagnostic} for any frozen encoder, alongside effective rank,
    participation ratio, condition number and channel/frame ablations
    (Sec.\ \ref{sec:analysis}).
    \item An \textbf{end-to-end supervised baseline} trained on the same
    encoder, providing an upper-bound anchor and an interesting
    "frozen-supervised" diagnostic.
\end{itemize}

\section{Setup}
\label{sec:setup}

\subsection{Dataset}
The \texttt{active\_matter} subset comprises 8\,750 / 1\,200 / 1\,300 sliding
windows of 16 frames at $224{\times}224$ spatial resolution and 11 physical
channels (concentration scalar, 2-channel velocity, 4-channel orientation
tensor, 4-channel strain-rate tensor); 45 unique $(\alpha,\zeta)$ combinations
total. Detailed statistics are in Tab.~\ref{tab:dataset} (Appendix).

\subsection{Encoder (shared between both SSL streams)}
We use a 3D ConvNeXt encoder with stages of dimensions
$(32, 64, 128, 256, 256)$ and per-stage block depths $(2, 2, 4, 8, 2)$, a
$2{\times}4{\times}4$ stem patch and 0.05 stochastic depth. The total spatial
downsampling is $32\times$ and temporal downsampling is $16\times$, yielding a
$(B, 256, 7, 7)$ bottleneck per clip. The encoder has 7.7\,M parameters.

\subsection{Evaluation protocol}
For every encoder we extract frozen features and run \emph{both}
(i) a single $\mathbf{nn.Linear}(256, 2)$ probe trained with MSE loss against
$z$-scored $(\alpha, \zeta)$ on the train split, and (ii) a $k$-Nearest
Neighbours regressor with hyperparameters swept over
$k\in\{1,3,5,10,20,50\}$, weights $\in\{$uniform, distance$\}$, metrics
$\in\{$cosine, euclidean$\}$ and feature normalization
$\in\{$z-score, $\ell_2$, z+$\ell_2\}$. Model selection uses the validation
split; the test split is touched only once per encoder for the final number.
Label normalization (z-score) is fit on \emph{train labels only}.

\section{SSL Objectives}
\label{sec:method}

\subsection{Latent prediction: SIGReg-JEPA (Person A)}
\label{sec:method-jepa}
\textcolor{red}{[Person A: fill in the SIGReg / LeJEPA section -- the
Epps--Pulley distributional regularizer formulated against $\mathcal{N}(0, I)$,
context/target temporal split, predictor architecture, etc.]}

\subsection{Pixel reconstruction: VideoMAE/SimMIM hybrid (Person B)}
\label{sec:method-pixel}
We adapt VideoMAE \citep{tong2022videomae} for our 3D ConvNeXt encoder. Three
adaptations are needed compared to the canonical ViT-based recipe:
\begin{enumerate}[topsep=2pt,itemsep=2pt]
    \item \textbf{Mask token replacement (SimMIM-style).} Depthwise 3D
    convolutions require the full input tensor; we cannot use VideoMAE's
    asymmetric encoder-decoder which drops masked tokens. Following SimMIM
    \citep{xie2022simmim} we instead replace each masked tube with a learnable
    per-channel mask token, broadcast over the tube extent, and feed the full
    tensor to the encoder.
    \item \textbf{Tube size aligned with encoder downsampling.} Per SimMIM
    Sec.~4.1.2, the optimal mask patch size matches the encoder stride. With
    $32\times$ spatial and $16\times$ temporal downsampling, our tube size is
    $(16, 32, 32)$, giving a $7{\times}7=49$ tube grid per sample.
    \item \textbf{Lightweight prediction head.} SimMIM Tab.~2 shows lightweight
    heads transfer better than heavy decoders. We use a 5-stage
    transposed-conv decoder (350\,k parameters) that mirrors the encoder
    downsampling, mapping $(256, 7, 7) \to (11, 16, 224, 224)$. Total model
    size is 8.07\,M parameters, well under the assignment's 100\,M cap.
\end{enumerate}

The reconstruction target is per-tube z-scored pixel values (VideoMAE
Sec.~3.3), and the loss is MSE over masked positions only (SimMIM Tab.~4
shows prediction-only beats full reconstruction):
\begin{equation}
\mathcal{L}_{\text{MAE}} = \frac{1}{|\mathcal{M}|}
    \sum_{(c,t,h,w) \in \mathcal{M}}
    \left( \hat{x}_{c,t,h,w} - \tilde{x}_{c,t,h,w} \right)^2,
\end{equation}
where $\mathcal{M}$ is the set of masked pixel coordinates and $\tilde{x}$ is
the per-tube z-scored target. The encoder is trained for 25 epochs with AdamW
($\text{lr}=1.5{\times}10^{-4}$, weight decay 0.05), a cosine schedule with
2-epoch linear warmup, and BF16 autocast.

\subsection{Supervised baseline (Person B)}
\label{sec:method-supervised}
We also train an end-to-end supervised baseline on the \emph{same} 3D ConvNeXt
encoder followed by an $\mathbf{nn.Linear}(256, 2)$ head, optimizing $z$-scored
MSE directly against $(\alpha, \zeta)$. This serves as an upper-bound anchor
and exposes how much information about the physical parameters is recoverable
from the encoder family at all. Trained for 30 epochs with AdamW
($\text{lr}=3{\times}10^{-4}$).

\section{Results}
\label{sec:results}

Tab.~\ref{tab:main-results} reports the headline numbers. Person~A's
SIGReg-JEPA row is left as a placeholder until their training completes; all
other rows are populated from the live \texttt{aggregated\_results.json}
artifact at the time of writing. We additionally include
\citet{qu2026spatiotemporal}'s published numbers as an external reference; note
that those values are obtained with a 100-epoch attentive-probe fine-tuning
protocol that the assignment explicitly forbids us to use, so they are not
directly comparable to our frozen-linear+kNN results -- they are listed for
context only.

\begin{table*}[t]
\centering
\caption{Active matter parameter prediction. Z-scored MSE (lower is better) on
the test split with a frozen encoder, evaluated with both a single linear probe
and $k$-NN regression. Effective rank and Epps--Pulley distance are computed on
the validation embedding set. We report mean$\pm$std across seeds where
available. Best per column among \emph{our} entries in \textbf{bold}.}
\label{tab:main-results}
\small
\begin{tabular}{lrrrrrrr}
\toprule
\textbf{Encoder} & \textbf{Params (M)} & \textbf{Linear} & \textbf{Linear $\alpha$} & \textbf{Linear $\zeta$} & \textbf{kNN} & \textbf{Eff.\ rank} & \textbf{Epps--Pulley} \\
\midrule
\textit{Ours:} & & & & & & & \\
\quad Supervised, 3 seeds (42, 43, 44)                & 7.72 & \textbf{0.106$\pm$0.006} & \textbf{0.029$\pm$0.006} & \textbf{0.182$\pm$0.017} & \textbf{0.106$\pm$0.020} & 3.55 & 0.086 \\
\quad VideoMAE-mask$0.60$, 25\,ep, 3 seeds              & 8.07 & 0.317$\pm$0.025 & 0.124$\pm$0.014 & 0.510$\pm$0.038 & 0.562$\pm$0.035 & 8.09 & 0.043 \\
\quad VideoMAE-mask$0.60$, 50\,ep, 1 seed (s$=46$)      & 8.07 & 0.338 & 0.179 & 0.497 & 0.389 & 6.47 & 0.050 \\
\quad VideoMAE-mask$0.75$, 1 seed                       & 8.07 & 0.299 & 0.075 & 0.522 & 0.672 & \textbf{8.87} & \textbf{0.039} \\
\quad VideoMAE-mask$0.90$, 2 seeds (44, 45)             & 8.07 & 0.255$\pm$0.002 & 0.108$\pm$0.003 & 0.402$\pm$0.007 & 0.503$\pm$0.039 & 7.92 & 0.042 \\
\quad SIGReg-JEPA (Person A)                            & 7.72 & TBD   & TBD   & TBD   & TBD   & TBD            & TBD            \\
\midrule
\multicolumn{8}{l}{\emph{External references (Qu et al., attentive-probe FT for 100 epochs -- different protocol):}} \\
\quad JEPA, ConvNeXt                        & ${\sim}3.3$ & 0.079 & --- & --- & --- & --- & --- \\
\quad VideoMAE, ViT-tiny/16                 & ${\sim}5.5$ & 0.160 & --- & --- & --- & --- & --- \\
\bottomrule
\end{tabular}
\end{table*}

Three observations stand out from Tab.~\ref{tab:main-results}.

\paragraph{(1) Isotropy / rank inversion.} The VideoMAE encoders have the
lowest Epps--Pulley distance to $\mathcal{N}(0,I)$ ($0.039$--$0.042$ vs.\
supervised $\sim$$0.087$) and the highest effective rank ($7.6$--$8.9$ vs.\
supervised $\sim$$3.5$), yet they have $\sim$$2.4$--$3.4{\times}$ \emph{worse}
linear probe MSE and $\sim$$3.8$--$8.6{\times}$ worse kNN MSE than the
supervised baseline. This is a direct quantitative confirmation of
\citet{qu2026spatiotemporal}'s claim that pixel reconstruction preserves
features that are physically uninformative; isotropy alone is not sufficient
for low downstream MSE when the encoder allocates rank to target-irrelevant
signal. We diagnose \emph{which} signal is target-irrelevant in
Sec.~\ref{sec:analysis} via channel-importance ablation.

\paragraph{(2) Mask-ratio sweep.} Increasing the mask ratio along $0.60 \to
0.75 \to 0.90$ \emph{monotonically} improves the linear probe MSE ($0.317 \to
0.299 \to 0.255$, two seeds at the endpoints). The kNN MSE is non-monotonic
but mask$=0.90$ is best on kNN among the 25-epoch runs ($0.50$, vs.\ $0.56$
at mask$=0.60$). This replicates Tong et al.'s finding that high mask ratios
($\sim$$90\%$) are optimal for video, extended here to physical fields with
$11$ channels of structured signal. Crucially, even the best VideoMAE
configuration (mask$=0.90$, std $0.002$ on linear MSE between seeds 44 and 45)
is still $\sim$$2.4{\times}$ worse than the supervised baseline -- the SSL
paradigm itself is the limiting factor, not the mask ratio.

\paragraph{(2b) Reconstruction loss is decoupled from downstream MSE.}
Training VideoMAE-mask$0.60$ for $50$ epochs instead of $25$ reduces the
validation reconstruction loss by $24\%$ ($0.501 \to 0.378$) but
\emph{does not improve the linear probe test MSE} (in fact $0.331 \to 0.338$,
slightly worse), although it does improve kNN test MSE ($0.610 \to 0.389$).
This empirically confirms the SimMIM observation that "stronger inpainting
capability does not necessarily lead to better fine-tuning performance"
(\citealp{xie2022simmim}, Tab.~2) on physical fields. The kNN improvement
without linear-probe improvement is itself instructive: longer SSL training
distributes more information across the embedding (longer training increases
the average per-dim std), which kNN can exploit non-linearly via local
neighbourhoods, while a linear probe still cannot find a useful low-dimensional
projection.

\paragraph{(3) Seed stability.} The supervised baseline is highly reproducible
(linear MSE $0.106$ and $0.098$ across two seeds, std $\sim$$0.005$, CKA
$0.97$). The VideoMAE-mask$0.60$ run is also stable across seeds (linear MSE
$0.331$ and $0.282$; std $\sim$$0.024$; CKA $0.96$), with the supervised
baseline showing $\sim$$5{\times}$ tighter seed-to-seed variation than VideoMAE
-- another consequence of supervised collapsing onto a low-rank
target-aligned subspace which is intrinsically less degenerate. The CKA
matrix (Fig.~\ref{fig:cka}) shows the four VideoMAE encoders cluster at CKA
$0.94$--$0.98$ regardless of mask ratio or seed, the two supervised seeds at
CKA $0.97$, and supervised-vs.-VideoMAE at CKA $0.70$--$0.75$ -- the SSL
objective is the dominant representation-shaping variable, while seed and
mask-ratio are secondary.

\section{Analysis}
\label{sec:analysis}

We evaluate every saved encoder with a representation-quality diagnostic suite
that captures collapse, isotropy and rank phenomena. All metrics operate on the
frozen-encoder embedding dump produced by \texttt{export\_embeddings.py}.

\paragraph{Collapse via per-dim std.} The sorted per-dimension standard
deviation curve (Fig.~\ref{fig:collapse}) makes representation collapse visible:
a collapsed encoder exhibits a heavy long tail with most dimensions near zero,
while a healthy encoder shows roughly uniform std.

\paragraph{Effective rank and participation ratio.} We compute
$\exp(H(p))$ where $p$ are the normalized eigenvalues of the embedding
covariance, plus the participation ratio
$(\sum \lambda_i)^2 / \sum \lambda_i^2$. Both are widely used proxies for the
true dimensionality of a representation manifold.

\paragraph{Condition number and off-diagonal correlation.} The covariance
condition number $\lambda_{\max} / \lambda_{\min}$ and the mean absolute
off-diagonal correlation reveal feature redundancy and ill-conditioning that
hurts both linear probes and kNN with Euclidean distances.

\paragraph{Epps--Pulley distance to $\mathcal{N}(0, I)$.} We repurpose the
LeJEPA SIGReg statistic \citep{balestriero2026lejepa} as a generic diagnostic:
the average squared characteristic-function error along random projections,
weighted by $e^{-t^2/2}$ and integrated to $t=3$. This is precisely Person A's
training objective (for SIGReg-JEPA), but applied here to \emph{any} encoder's
embeddings as a goodness-of-fit-to-isotropic-Gaussian metric.

\paragraph{Channel-importance ablation.} For each of the 11 input channels we
zero that channel at inference time, re-export embeddings on all splits, refit
the linear probe and measure the test-MSE delta. The most informative finding
is the \emph{difference} between supervised and VideoMAE: zeroing the
concentration scalar field hurts supervised by only $\Delta$MSE $+0.029$ but
hurts VideoMAE-mask$0.60$ by $\Delta$MSE $+0.264$ -- nearly an order of
magnitude more reliance. This is direct evidence that VideoMAE's pixel
reconstruction loss disproportionately weights the dense scalar concentration
field, which is largely uninformative for $(\alpha, \zeta)$. The supervised
encoder, by contrast, distributes channel reliance more evenly across the
orientation tensor (orient\_yy, $\Delta$MSE $+0.051$), velocity (vel\_x,
$+0.036$) and concentration ($+0.029$). Full per-channel deltas are in
Tab.~\ref{tab:channel-ablation} (Appendix).

\paragraph{Frame-budget ablation.} Re-exporting embeddings with the encoder
fed only the first $T \in \{4, 8, 12, 16\}$ frames (latter frames replaced by
the last available frame) shows that supervised saturates around $T=8$ frames
($\text{MSE} = 0.128$) while VideoMAE is more sample-hungry temporally:
mask$0.60$ improves from $0.462 \to 0.247$ between $T=4$ and $T=12$, and
mask$0.75$ from $0.542 \to 0.297 \to 0.259$. VideoMAE benefits substantially
from full temporal context, suggesting the masked-tube objective relies on
genuine temporal information for reconstruction.

\paragraph{Top-PC alignment: how many leading directions encode $(\alpha,
\zeta)$?} For each encoder we project the (z-scored) frozen embeddings onto
their top-$K$ principal components and fit a closed-form ridge regression
linear probe on just those $K$ features (no SGD); test MSE as $K$ grows tells
us how concentrated the target-relevant signal is at the top of each
encoder's spectrum. The result (Fig.~\ref{fig:pc-alignment}) is striking:
\emph{supervised reaches its full performance at $K=2$} (test MSE $0.083$ at
$K=2$ vs.\ $0.086$ at $K=174$), while \emph{VideoMAE needs essentially the
entire 256-d embedding to reach its (worse) plateau} (mask$0.60$: $0.55 \to
0.25$ from $K=2$ to $K=174$, a $2.2{\times}$ drop). This is the
single sharpest demonstration of the rank-vs.-alignment trade-off in our
report: the supervised encoder's leading $2$ principal directions are
essentially the $\alpha$ and $\zeta$ directions, while VideoMAE's leading
directions encode something else entirely and the target signal is smeared
across the full spectrum.

\paragraph{Per-class separability: how distinguishable are the 45 unique
$(\alpha, \zeta)$ combinations in embedding space?} The dataset has exactly
$45 = 5 \times 9$ unique parameter combinations. We compute each combination's
centroid in z-scored embedding space on the train split and report (a) the
mean within-class distance to the centroid, (b) the mean pairwise
between-centroid distance, and (c) their ratio (a Fisher-style separability
score). We also report a nearest-centroid classification accuracy on test as
a diagnostic (random baseline $\approx 2.2\%$).

\begin{itemize}[topsep=2pt,itemsep=2pt]
    \item Supervised has Fisher separability $3.4$ vs.\ VideoMAE's
    $1.3$ (a $2.6{\times}$ gap), driven by a $2{\times}$ smaller within-class
    spread \emph{and} a $\sim$$30\%$ larger between-centroid distance.
    \item Nearest-centroid test accuracy is $42\%$ for supervised vs.\ $4\%$
    for VideoMAE -- supervised correctly classifies its target combinations
    \emph{ten times more often} than VideoMAE despite the simpler 1-NN
    decision rule.
    \item Across all VideoMAE configurations (mask $0.60/0.75/0.90$, two
    seeds) the Fisher score sits in a tight band $1.29$--$1.35$, again showing
    that the SSL objective -- not the mask ratio -- is the variable that
    matters.
\end{itemize}

\paragraph{Per-channel reconstruction MSE.} As a direct test of \emph{which
output channels} VideoMAE actually tries to reconstruct, we feed validation
clips through each VideoMAE encoder+decoder and measure the per-channel MSE
on masked positions (in the per-tube z-scored target space). Across all
VideoMAE configurations (mask $0.60/0.75/0.90$, two seeds), the result is
remarkably consistent:
\emph{the concentration scalar field is reconstructed to MSE $\approx 0.22$,
while the four orientation tensor channels and four strain tensor channels
all sit in the $0.55$--$0.78$ range, $2.5$--$3.5{\times}$ harder} (Tab.
\ref{tab:recon-channel-mse}, Appendix). The 2-channel velocity field is
intermediate ($0.41$--$0.57$). This is a \emph{causal} confirmation of the
channel-importance ablation: VideoMAE allocates decoder capacity to the
concentration field because it can be reconstructed sharply, then relies on
that channel at downstream evaluation time -- exactly the failure mode the
channel-zeroing experiment exposed. Increasing the mask ratio from $0.60$ to
$0.90$ uniformly raises reconstruction difficulty (mean per-channel MSE
$0.51 \to 0.55 \to 0.68$) but does not change \emph{which} channel is easiest.

Together, the channel-importance, frame-budget, top-PC-alignment,
per-class-separability, and per-channel-reconstruction analyses give five
\emph{independent} quantitative characterizations of the same underlying
phenomenon: VideoMAE's pixel reconstruction objective allocates
representational capacity to the concentration scalar field and to high-rank
but target-misaligned directions, producing an embedding space where the 45
governing parameter combinations are not cleanly separated. Examples of
input / masked / reconstructed / residual visualizations for each VideoMAE
mask ratio are in Fig.\ \ref{fig:recon-grids} (Appendix).

\section{Limitations and Ethics}

Our headline numbers use 2 seeds for supervised and 2 seeds for VideoMAE
mask$=0.60/0.90$, plus one seed for the mask$=0.75$ ablation; the agreement
across seeds (CKA $\geq 0.96$, std $\leq 0.034$ on linear MSE) is tight, but
3+ seeds would let us report formal confidence intervals. We did not train a
from-scratch ViT-based VideoMAE that would precisely replicate
\citet{qu2026spatiotemporal}'s baseline -- our VideoMAE deliberately uses the
colleague's ConvNeXt encoder for causal cleanliness of the SSL-objective
comparison, at the cost of differing slightly from the published recipe.
Channel-importance and frame-budget ablations re-export embeddings under a
zeroed input or truncated time axis; this measures the encoder's reliance on
those inputs at \emph{evaluation} time and not the role of those inputs during
SSL training (an interesting future direction). The dataset is a synthetic
fluid simulation; no human or animal data are involved.

\paragraph{Compute accounting.} All experiments ran on 2$\times$ NVIDIA B200
180\,GB GPUs (PyTorch 2.11.0+cu128, BF16 autocast) on a single rented node.
Per training run: VideoMAE main 25 epochs at ${\sim}172$\,s/epoch
(${\sim}1.2$\,h); supervised 30 epochs at ${\sim}90$\,s/epoch (${\sim}45$\,min);
mask$=0.75$ and mask$=0.90$ ablations 12 epochs warm-started
(${\sim}35$\,min each); a second seed for both VideoMAE-mask$0.60$ and
supervised. Per encoder: linear-probe sweep over $36$ trials and kNN sweep
over $72$ trials (${\sim}10$\,min each), the channel-importance ablation
($11$ channels $\times$ $3$ splits export plus probe per channel,
${\sim}5$\,min), and the frame-budget ablation ($4$ frame budgets $\times$ 3
splits, ${\sim}3$\,min). Total compute: ${\sim}10$ GPU-hours of training plus
${\sim}5$ GPU-hours of evaluation, well within the project budget. Peak VRAM
during training was ${\sim}71$\,GB on the VideoMAE GPU and ${\sim}51$\,GB on
the supervised GPU at batch size $64$. We hit the system OOM killer once early
on with $12{+}8$ DataLoader workers running concurrently with a
dataset-hash computation; reducing to $6{+}4$ workers resolved it without
losing any training progress thanks to per-epoch checkpointing.

\section*{Statement of Contributions}

\textbf{Person A} authored the \texttt{physrl/} package (3D ConvNeXt encoder,
JEPA latent-prediction objective, VICReg and SIGReg loss functions, the
training/eval/sweep CLIs reused unchanged here, and the related-work review of
JEPA-style SSL). Person~A's stream covers Sec.~\ref{sec:method-jepa} and the
SIGReg-JEPA rows of Tab.~\ref{tab:main-results}.

\textbf{Person B} authored the \texttt{videomae/} package (the
VideoMAE/SimMIM-hybrid trainer, end-to-end supervised baseline,
representation-quality analysis suite, channel-importance / frame-budget /
top-PC-alignment / per-class-separability / per-channel-reconstruction
analyses, CKA cross-encoder comparison, reconstruction-grid visualizations,
the figure regeneration script, the detached training launchers, the
dataset hash and ENV.md/requirements lock for reproducibility) and the
corresponding paper sections
(\ref{sec:method-pixel},~\ref{sec:method-supervised},~\ref{sec:analysis}).
Files in \texttt{videomae/} whose names also appear in \texttt{physrl/} are
derivative copies with rewritten imports; modifications are documented in their
module docstrings.

\section*{Appendix}

\paragraph{Dataset statistics (Tab.~\ref{tab:dataset}).}
\begin{table}[h]
\centering
\caption{The \texttt{active\_matter} subset of \emph{The Well} \citep{ohana2025well} as we use it.}
\label{tab:dataset}
\small
\begin{tabular}{lrrr}
\toprule
\textbf{Split} & \textbf{HDF5 files} & \textbf{Trajectories} & \textbf{Sliding windows (stride 1)} \\
\midrule
Train & 45 & 175 & 8\,750 \\
Valid & 16 &  24 & 1\,200 \\
Test  & 21 &  26 & 1\,300 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Channel-importance deltas (Tab.~\ref{tab:channel-ablation}).}
$\Delta$test-MSE when each input channel is zeroed at inference time, computed
on the test split with a frozen-encoder linear probe. Higher $\Delta$ = more
reliance on that channel.

\begin{table*}[h]
\centering
\caption{Per-channel ablation $\Delta$test-MSE (z-scored). Bold = top channel for that encoder.}
\label{tab:channel-ablation}
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lrrrrrrrrrrr}
\toprule
\textbf{Encoder} & \textbf{conc.} & \textbf{vel\_x} & \textbf{vel\_y} & \textbf{or\_xx} & \textbf{or\_xy} & \textbf{or\_yx} & \textbf{or\_yy} & \textbf{st\_xx} & \textbf{st\_xy} & \textbf{st\_yx} & \textbf{st\_yy} \\
\midrule
Supervised      & $+0.029$ & $+0.036$ & $+0.028$ & $+0.024$ & $-0.017$ & $-0.011$ & $\mathbf{+0.051}$ & $+0.044$ & $+0.012$ & $+0.028$ & $+0.023$ \\
VideoMAE-m$0.60$ & $\mathbf{+0.264}$ & $+0.058$ & $+0.030$ & $+0.112$ & $-0.024$ & $-0.005$ & $+0.001$ & $+0.195$ & $+0.025$ & $+0.077$ & $+0.040$ \\
VideoMAE-m$0.75$ & $+0.228$ & $+0.113$ & $+0.034$ & $+0.215$ & $-0.024$ & $+0.005$ & $+0.111$ & $\mathbf{+0.281}$ & $+0.024$ & $+0.105$ & $+0.058$ \\
\bottomrule
\end{tabular}
\end{table*}

\paragraph{Per-channel reconstruction MSE (Tab.~\ref{tab:recon-channel-mse}).}
Mean over $3$ random masking realizations on the validation split, in the
per-tube z-scored target space. Lower is "easier to reconstruct."

\begin{table*}[h]
\centering
\caption{Per-channel masked-position reconstruction MSE for each VideoMAE
configuration. The concentration scalar field is reconstructed
$\sim$$2.5{\times}$ better than orientation/strain tensors.}
\label{tab:recon-channel-mse}
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lrrrrrrrrrrr}
\toprule
\textbf{Encoder} & \textbf{conc.} & \textbf{vel\_x} & \textbf{vel\_y} & \textbf{or\_xx} & \textbf{or\_xy} & \textbf{or\_yx} & \textbf{or\_yy} & \textbf{st\_xx} & \textbf{st\_xy} & \textbf{st\_yx} & \textbf{st\_yy} \\
\midrule
VideoMAE-m$0.60$    & $\mathbf{0.218}$ & $0.483$ & $0.448$ & $0.575$ & $0.578$ & $0.562$ & $0.577$ & $0.570$ & $0.537$ & $0.545$ & $0.571$ \\
VideoMAE-m$0.60$ s$=43$ & $\mathbf{0.223}$ & $0.509$ & $0.454$ & $0.605$ & $0.563$ & $0.563$ & $0.594$ & $0.577$ & $0.546$ & $0.554$ & $0.594$ \\
VideoMAE-m$0.75$    & $\mathbf{0.217}$ & $0.430$ & $0.406$ & $0.625$ & $0.622$ & $0.619$ & $0.627$ & $0.629$ & $0.593$ & $0.597$ & $0.630$ \\
VideoMAE-m$0.90$    & $\mathbf{0.218}$ & $0.574$ & $0.561$ & $0.780$ & $0.765$ & $0.764$ & $0.781$ & $0.779$ & $0.741$ & $0.742$ & $0.779$ \\
\bottomrule
\end{tabular}
\end{table*}

\paragraph{Figures.}
The full PDF figures live in \texttt{videomae/artifacts/} and are rebuilt by
\texttt{videomae/plot\_from\_json.py}, \texttt{videomae/plot\_ablations.py},
\texttt{videomae/cka\_compare.py}, \texttt{videomae/pc\_alignment.py},
\texttt{videomae/visualize\_embeddings.py}, and
\texttt{videomae/visualize\_reconstructions.py}. Specifically:
\begin{itemize}[topsep=2pt,itemsep=0pt]
    \item Fig.~\ref{fig:collapse}: per-dim embedding std (sorted) for every
    encoder. \texttt{videomae/artifacts/figures/figures.pdf}, panel "Per-dim
    embedding std".
    \item Fig.~\ref{fig:cka}: CKA matrix between all encoders.
    \texttt{videomae/artifacts/cka/cka\_valid.pdf}.
    \item Fig.~\ref{fig:pc-alignment}: top-$K$ PC linear-probe MSE curve.
    \texttt{videomae/artifacts/pc\_alignment/pc\_alignment.pdf}.
    \item Fig.~\ref{fig:recon-grids}: per-channel input / masked /
    reconstructed / residual grids for VideoMAE-mask$0.60/0.75/0.90$.
    \texttt{videomae/artifacts/recon\_viz/recon\_*\_t8.pdf}.
\end{itemize}
For grading convenience, all four PDFs are committed to git on the
\texttt{videomae} branch.

% References
\begin{thebibliography}{9}

\bibitem[Qu et al.(2026)]{qu2026spatiotemporal}
Helen Qu, Rudy Morel, Michael McCabe, Alberto Bietti, Fran\c{c}ois Lanusse,
Shirley Ho, Yann LeCun.
\newblock Representation Learning for Spatiotemporal Physical Systems.
\newblock \emph{ICLR 2026 Workshop on AI \& PDE}, arXiv:2603.13227, 2026.

\bibitem[Tong et al.(2022)]{tong2022videomae}
Zhan Tong, Yibing Song, Jue Wang, Limin Wang.
\newblock VideoMAE: Masked Autoencoders are Data-Efficient Learners for
Self-Supervised Video Pre-Training.
\newblock \emph{NeurIPS 2022}, arXiv:2203.12602.

\bibitem[Xie et al.(2022)]{xie2022simmim}
Zhenda Xie, Zheng Zhang, Yue Cao, Yutong Lin, Jianmin Bao, Zhuliang Yao,
Qi Dai, Han Hu.
\newblock SimMIM: A Simple Framework for Masked Image Modeling.
\newblock \emph{CVPR 2022}, arXiv:2111.09886.

\bibitem[Balestriero \& LeCun(2026)]{balestriero2026lejepa}
Randall Balestriero, Yann LeCun.
\newblock LeJEPA: Provable and Scalable Self-Supervised Learning Without the
Heuristics.
\newblock arXiv:2511.08544, 2026.

\bibitem[Bardes et al.(2021)]{bardes2021vicreg}
Adrien Bardes, Jean Ponce, Yann LeCun.
\newblock VICReg: Variance-Invariance-Covariance Regularization for
Self-Supervised Learning.
\newblock arXiv:2105.04906, 2021.

\bibitem[Liu et al.(2022)]{liu2022convnext}
Zhuang Liu, Hanzi Mao, Chao-Yuan Wu, Christoph Feichtenhofer, Trevor Darrell,
Saining Xie.
\newblock A ConvNet for the 2020s.
\newblock \emph{CVPR 2022}.

\bibitem[Ohana et al.(2025)]{ohana2025well}
Ruben Ohana et al.
\newblock The Well: a Large-Scale Collection of Diverse Physics Simulations
for Machine Learning.
\newblock arXiv:2412.00568, 2025.

\bibitem[Balestriero et al.(2023)]{balestriero2023cookbook}
Randall Balestriero et al.
\newblock A Cookbook of Self-Supervised Learning.
\newblock arXiv:2304.12210, 2023.

\bibitem[He et al.(2022)]{he2022mae}
Kaiming He, Xinlei Chen, Saining Xie, Yanghao Li, Piotr Doll\'ar, Ross Girshick.
\newblock Masked Autoencoders Are Scalable Vision Learners.
\newblock \emph{CVPR 2022}, arXiv:2111.06377.

\end{thebibliography}

\end{document}
